---
title: "Checking if IPC AMN sample size requeriments were met"
author: "Tomás Zaba"
format: html
vignette: >
  %\VignetteIndexEntry{Checking if IPC AMN sample size requeriments were met}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: global-setup
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

library(ipccheckr)
```
Evidence on the prevalence of acute malnutrition used in the IPC Acute Malnutrition (IPC AMN) can come from different sources with the data collected in different ways:  representative surveys, screenings or community-based surveillance system (known as sentinel sites). The IPC set minimum sample size requirements for each of these sources. Details can be read from the [IPC Manual version 3.1 ](https://www.ipcinfo.org/ipcinfo-website/resources/ipc-manual/en/). 

In the IPC AMN analysis workflow, the very first step of a data analyst is to check if these requirements were met. This is done for each area meant to be included in the *de facto* IPC AMN analysis. For this, `mwana` provides a handy function: `check_sample_size()`.

To demonstrate its usage, we will use a built-in sample dataset `anthro.01`.

```{r}
#| label: view-data
#| echo: true
#| eval: false

head(anthro.01)
```

```{r}
#| label: view-data.
#| echo: false
#| eval: true
head(anthro.01)
```

`anthro.01` contains anthropometry data from SMART surveys from anonymized locations. We can check further details about the dataset by calling `help(anthro.01)` in `R` console. 

Now that we are acquainted with the dataset, we can now execute the task. To achieve this we simply do:

```{r}
#| label: check
#| echo: true
#| eval: false

check_sample_size(
  df = anthro.01,
  .group = cluster,
  .data_type = "survey"
)
```

Or we can also choose to chain the data object to the function using the pipe operator: 
```{r}
#| label: pipe_operator
#| echo: true
#| eval: false

anthro.01 |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

Either way, the returned output will be: 
```{r}
#| label: view_check
#| echo: false
#| eval: true

anthro.01 |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

A table (of class `tibble`) is returned with three columns:  

  + Column `groups` counts the number of unique cluster ID's in the dataset.  
  + Column `n_obs` counts the number of children the dataset.  
  + Column `meet_ipc` indicates whether the IPC AMN sample size requirements (for surveys in this case) were met or not

We often deal with multi-area dataset. We can get a summarised table by area as follows: 
```{r}
#| label: group_by
#| echo: true
#| eval: false

## Load dplyr package ----
library(dplyr)

## Use group_by() function ----
anthro.01 |> 
  group_by(area) |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

This will return: 

```{r}
#| label: view_group_by
#| echo: false
#| eval: true
#| message: false

## Load the `dplyr` library ----
library(dplyr)

## Use the `group_by()` function ----
anthro.01 |> 
  group_by(area) |> 
  check_sample_size(
    .group = cluster,
    .data_type = "survey"
  )
```

For screening or sentinel site-based data, we approach the task the same way; we only have to change the `.data_type` parameter to "screening" or to "ssite" as appropriate, as well as by supplying `.group` with the right column name of the sub-areas inside the main area (villages, localities, comunas, communities, etc) .
